---
title: "Week6"
author: "Johanne S. Rejsenhus"
date: "2025-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Good afternoon*  
Today we are going to work with some of the statistical concepts Anna went over yesterday. We will check for normality and implement correlation tests in R. The goal of this markdown is to play around with plotting, transforming and testing assumptions of normality and try to gain an understanding of the concepts. 


**Structure of the markdown:**  
Part 1:
- Visually inspect normality of data
- Test assumptions of normality
- Try standardization
- Perform correlation tests 

Part 2:
- Apply these techniques in a real data example


Start with the first exercise, and then continue in order. Feel free to work together, and see how far you can get.   

The important thing is to learn, not to necessarily do it all!
________________________________________________________________________________________


## Load the packages for today

```{r}
# Install pacman first (only once):
install.packages("pacman")

# Load and install packages:
pacman::p_load(ggplot2, dplyr, tidyr) #So it installs the library if needed + imports it in one step.

```

# Part 1:
For Part one we will walk through the necessary steps of making an analysis of cultural data (except making and fitting models - but that's for next week :)). 

_We would like to investigate whether the personal savings ratio is larger for countries with large disposable income_

## The dataset for the first part
We are gonna work with the df_lcs dataset which is a built-in classic dataset in R, containing economic and demographic indicators for 50 countries, covering the years 1950–1960.

```{r}

df_lcs <- as.data.frame(LifeCycleSavings)
head(df_lcs)

```
The dataset has 50 rows (countries) and 5 numeric variables:

sr – Aggregate personal savings ratio - Average % of disposable personal income that households save.
pop15 – % of population under age 15
pop75 – % of population over age 75
dpi – Real per-capita disposable income
ddpi – Growth rate of dpi


## Check for normality
- Histograms
- QQ-plots
- Statistical normality tests

First we plot the data to visually inspect how the data is distributed - we hope for normal distributions. 

We use 'ggplot' to plot in r and there is so many cool plot you can make to visualise your data. If you would like to try and spice things up, go to: https://r-graph-gallery.com and look around. 

```{r}
# Now we plot the histograms of the savings ratio and the disposable income 

# Savings Ratio
ggplot(df_lcs, aes(x = sr)) + # Here we specify the aesthetics (Which is "just" specifying the x and potentially the y axis.)
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightblue", color = "black") + # Here we specify the type of plot and that it should be a density plot. 
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Savings Ratio (sr)", x = "Savings Ratio", y = "Density") # And we give it a title and some axis names

#Disposable income
ggplot(df_lcs, aes(x = dpi)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Disposable Income (dpi)", x = "Disposable Income", y = "Density")

```

_Exercise_: Explain what you see, are the distributions normal, if not, describe how they look. 

**The Distribution of Savings Ratio looks rather normal, while the Distribution of Disposable Income doesn't look that normal as its skewed towards the left side (positive skew).**

## QQ plots
The QQ-plot stands for Quantile-Quantile plot. Is plots the theoretical quantiles on the x-axis. That is the quantiles the data would have if it was perfectly normal, and then it plots the actual quantiles on the y-axis. 

If you forgot, quantiles are cut points that divide your data into equal-sized groups. 

```{r}
# Now we plot the QQ-plots of the savings ratio and the disposable income 

ggplot(df_lcs, aes(sample = sr)) + #For qq-plots we specify a 'sample' instead of an x or y axis
  stat_qq(color = "blue") + # Colours 
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Savings Ratio (sr)")

ggplot(df_lcs, aes(sample = dpi)) +
  stat_qq(color = "darkgreen") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Disposable Income (dpi)")

```

_Exercise_: Does this fit with what you observed before when you plotted the histograms? Explain why.

**Yes, it also follows what was observed above, as it shows the (sr) following the line nicely, while the (dpi) deviates a lot from the line.**


## Test for normality - Shapiro-wilk test

The Shapiro-Wilk test is a statistical test used to check whether a variable is approximately normally distributed. It is one of the most common tests for normality in R. It test whether there is strong enough evidence to reject the null hypothesis, which is that the data is normal. 

How It Works:
- Null hypothesis (H₀): The data are normally distributed.
- Alternative hypothesis (H₁): The data are not normally distributed.

The test calculates a W statistic that measures how well the data’s distribution matches a normal distribution.


```{r}

shapiro.test(df_lcs$sr)
shapiro.test(df_lcs$dpi)

```

How to interpret the results:
Look at the p-value:
- p > 0.05 → Fail to reject H₀ → Data can be treated as approximately normal.
- p ≤ 0.05 → Reject H₀ → Consider non-parametric methods or transformations.

W statistic:
- W ranges from 0 to 1.
- Values closer to 1 indicate data closer to normal.

_Exercise_: Report the results for both variables, are they normal and why/why not?

**For (sr) the p-value = 0.5836 (p > 0.05), which mean the data can be treated as normal. However, for (dpi) the p-value = 0.0000328 (p < 0.05), which mean the data shouldn't be treated as normal.**


## Transformation - Lets see if we can force the dpi-variable to become normal. 
We will use log() (logarithmic) and sqrt() (square root) transformation methods and see if it helps.

```{r}

# Log transformation
df_lcs$log_dpi <- log(df_lcs$dpi)

# Sqrt transformation
df_lcs$sqrt_dpi <- sqrt(df_lcs$dpi)

#plot it 
ggplot(df_lcs, aes(x = log_dpi)) +
  geom_histogram(aes(y = ..density..), bins = 15, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Disposable Income (log)", x = "Disposable Income", y = "Density")

ggplot(df_lcs, aes(x = sqrt_dpi)) +
  geom_histogram(aes(y = ..density..), bins = 15, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Disposable Income (sqrt)", x = "Disposable Income", y = "Density")


# test both
shapiro.test(df_lcs$log_dpi)
shapiro.test(df_lcs$sqrt_dpi)

```
_Excercise_: Does any of the transformations make the dpi variable normal?

**Looking at the histograms, the data seems to be less skewed with the log transformation showing a more normal looking distribution. For the log transformed data, the shapiro-wilk test showed a p-value of 0.04, while the squareroot transformation showed a p-value of 0.003. Thus, the log transformation seems to be a bit more normal, as it's closer to 1, but still not considered statistically normal.**


## Standardization 
For this simple correlation analysis, we will skip it, because the results are already interpretable. However, if we wanted to fit a statistical model (e.g., regression) to this data, standardization would be useful. This is because the model coefficients would then represent changes in standard deviations, rather than raw units, making them easier to compare across variables.

This is a method that you will not have to use often, but it is important to know. 

```{r}

# This is how you standardize the data (transform to z-scores)
scaled_data <- scale(df_lcs)
head(scaled_data)

```


_Exercise_: Is standardized data easy to directly interpret? why/why not? (what is a unit now?)

**It's perhaps a bit more unintuitive, as they are not directly interpretable. But it makes it easier to interpret variables with each other that were in different units before.**
 
## Correlation tests
Now, We unfortunately do not have normally distributed data so we should use the spearman correlation test. However, for teaching purposes you will do both and report them to compare the results. 

Pearson:
Measures: Linear relationship
Assumptions: Variables approximately normal, linear, no extreme outliers
Use when: Data are normal and linear

Spearman: 
Measures: Monotonic relationship (rank-based)
Assumptions: None
Use when: Data are skewed, contain outliers, or relationship is monotonic but non-linear

```{r}

# Pearsons correlation - Assumes normality 

cor.test(df_lcs$sr, df_lcs$dpi, method = "pearson")

```
_Excercise_: Report on the results. Whats the confidence interval, the p-value and the correlation and how should they be interpreted?

**(sr) was found to not be significantly correlated with (dpi), r(48) = .22, 95% CI [-0.06, 0.47], p = .12**

**Once the p-value is > 0.05, there's no statistical significance, and the closer the correlation is to 1, the better is the relationship between the variables. In this case, (dpi) only explains 22% of the variance in (sr).**

**The confidence interval is [-0.06177804, 0.46988555], showing that we are 95% sure of the true mean lying within this interval.**

You can find the APA format of reporting here: https://www.statology.org/how-to-report-pearson-correlation/ :)


```{r}
# Spearman correlation - does not assume normality

cor.test(df_lcs$sr, df_lcs$dpi, method = "spearman")

```

_Excercise_: Report on the results. Whats the rho and the p-value and how should they be interpreted?

**(sr) was found to now be significantly correlated with (dpi), r(48) = .28, p = .04**

**Once the p-value is < 0.05, there's a statistical significance, and the the correlation is a bit closer to 1 now.**

You can find the APA format of reporting Spearmans correlation here: https://www.statology.org/how-to-report-spearman-correlation/ :) 


Okay, so why do we have different result?

_Exercise_: What causes the difference in the results?

**The difference is that we now used the right model on our non-parametric data that is more well suited when dealing with data that doesn't follow a normal distribution.**

Well done!! now for part 2


# Part 2:
## Now we try and use our new knowledge for a mini-project. For this I have made a checklist you can run through

### Step 1: Understand Your Data
- [ ] Load the dataset into R  
- [ ] Identify which variables are relevant for your analysis  

### Step 2: Explore Distributions
- [ ] Plot density histograms for the selected variables - or maybe try some other plots (violinplots, boxplots etc.)
- [ ] Create Q-Q plots to visually check normality  
- [ ] Run a Shapiro-Wilk test for normality

- Perhaps do a transformation here if needed

### Step 3: Standardize Variables (if needed)
- [ ] Apply `scale()` to standardize if comparing across variables  

### Step 4: Check Assumptions for Correlation
- [ ] If variables look normally distributed → Pearson correlation  
- [ ] If variables are not normal or have outliers → Spearman correlation  

### Step 5: Run Correlation Tests
- [ ] Compute Pearson correlation with `cor.test(x, y, method="pearson")`  
- [ ] Compute Spearman correlation with `cor.test(x, y, method="spearman")`  
- [ ] Interpret results  

### Step 6: Interpret Results
- [ ] Report the correlation coefficient (strength & direction of relationship)  
- [ ] Report the p-value (evidence for/against significant relationship)  
- [ ] Relate findings back to the research question / context of the dataset  


You now have a lot of freedom to find a good question to test, but first lets take a look at the data :)

```{r }

df_states <- as.data.frame(state.x77)
head(df_states)

```

What dataframe are you looking at, make a note of the column names and their meaning and discuss with your group what an interesting thing to investigate could be. 

Few points of focus:
- Try to set up your analysis like part one 
- Remember to comment your code, so you understand whats happening. 
- Explain whats happening along the way and report on your results

Now the floor is yours ...

**I will look at life expectancy and murder across states**

```{r}

# First I'll plot the histogram for the two variables: 

# Savings Ratio
ggplot(df_states, aes(x = `Life Exp`)) + # Here we specify the aesthetics (Which is "just" specifying the x and potentially the y axis.)
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightblue", color = "black") + # Here we specify the type of plot and that it should be a density plot. 
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Life Expectancy", x = "Life Expectancy in Years", y = "Density") # And we give it a title and some axis names

#Disposable income
ggplot(df_states, aes(x = Murder)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Murders", x = "Murder per 100k Capita", y = "Density")

```
**Looking at the histograms, life expectancy looks to be slightly more normal.**


```{r}

# Then we'll look at the qq-plots: 

ggplot(df_states, aes(sample = `Life Exp`)) + #For qq-plots we specify a 'sample' instead of an x or y axis
  stat_qq(color = "blue") + # Colours 
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Life Expectancy")

ggplot(df_states, aes(sample = Murder)) +
  stat_qq(color = "darkgreen") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Murder")

```
**Both variables have nice Q-Q-plots with life expectancy showing a bit less deviation from the line.**

```{r}

# Then we'll do the shapiro wilk test: 

shapiro.test(df_states$`Life Exp`)
shapiro.test(df_states$Murder)

```
**For life expectancy, p = 0.44 (p > 0.05), so the data can be treated as approximately normal.**

**For murder, p = 0.04 (p < 0.05), thus, the data should not be treated as normal.**

```{r}

# Thus, we'll try to transform our murder data set: 

# Log transformation
df_states$log_murder <- log(df_states$Murder)

# Sqrt transformation
df_states$sqrt_murder <- sqrt(df_states$Murder)

#plot it 
ggplot(df_states, aes(x = log_murder)) +
  geom_histogram(aes(y = ..density..), bins = 15, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Murder (log)", x = "Murder per 100k Capita (log)", y = "Density")

ggplot(df_states, aes(x = sqrt_murder)) +
  geom_histogram(aes(y = ..density..), bins = 15, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Murder (sqrt)", x = "Murder per 100k Capita (sqrt)", y = "Density")

# test both
shapiro.test(df_states$log_murder)
shapiro.test(df_states$sqrt_murder)

```
**It shows that the sqrt transformed data can be treated as approximately normal (p > 0.05), which is also conveyed through the histogram.**

```{r}

# Because we're comparing data across variables, we should remember to standardize it. 

scaled_data_states <- scale(df_states)

head(scaled_data_states)

```

```{r}

# Now we can apply Pearson's correlation, as the variables look normally distributed: 

cor.test(df_states$`Life Exp`, df_states$sqrt_murder, method = "pearson")

```

Interpret and: 
- [ ] Report the correlation coefficient (strength & direction of relationship)  
- [ ] Report the p-value (evidence for/against significant relationship)  
- [ ] Relate findings back to the research question / context of the dataset  

**Life Expectancy was found to be significantly correlated with Murder, r(48) = -0.77, 95% CI [-0.86, -0.62], p = .0.00000000006686**

**Once the p-value is < 0.05, there's statistical significance, and the closer the correlation is to 1, the better is the relationship between the variables, though it's a negative relationship.**

**The two variables are negatively correlated, so the higher the life expectancy is, the less murders are there.**


















