---
title: "Week5"
author: "Johanne S. Rejsenhus"
date: "2025-09-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Good afternoon*  
Today we are going to work with in a R markdown file for the first time. It works just like a Python notebook with cell-blocks of r code and in between you can write (like I am here). 
The goal of this markdown is that you have tried to translate some of the exercises you already have done in python. We are doing this so you can get your hands dirty with some syntax translation and hopefully get a bit more comfortable in r. 


**Structure of the markdown:**  
- Loading packages
- Variables
- Container variables
- Accessing elements
- Conditional statements
- Loops 
- Functions
- Data organisation


Start with the first exercise, and then continue in order. Feel free to work together, and see how far you can get.   
The important thing is to learn, not to necessarily do it all!
________________________________________________________________________________________


## Loading packages
Instead of installing and importing libraries in r seperately, we can do it in one go using the library pacman. It has a function p_load that allows for importing multiple libraries in one go! 

```{r Loading packages}

# Install pacman first (only once):
# install.packages("pacman")

# Load and install packages:
pacman::p_load(ggplot2, dplyr, tidyr) #So it installs the library if needed + imports it in one step.

```


## Variables 
Add some values to the variables below and test if they are the right variable type.  

Remember that in R, assignment is usually done with <- instead of a = as it is in python. 
Strings are written with quotation marks (" ").
Boolean values are written as TRUE or FALSE (all caps).
You can use the class() function in R to check the data type of a variable.


```{r}

# Adding values
my_string <- "Hello World!!"
my_integer <- 42
my_boolean <- TRUE
my_float <- 42.21

# Testing the class
class(my_string)
class(my_integer)
class(my_boolean)
class(my_float)

```


## Container variables
In R we have slight differences in the container types. 

In Python, we have lists, the equivalent in R is also called a list, created with the list() function.  

In Python, dictionaries store key–value pairs. In R, there is no direct equivalent, but you can achieve something similar with a named list (created with list(name = value)), or in some cases with environments or data frames, depending on the use case.

In Python, tuples are ordered collections that cannot be changed (immutable). In R, there is no immutable equivalent. The closest structure is a vector (created with c()), which is ordered but still mutable.

In Python, sets are unordered collections of unique items. In R, there is no dedicated set type, but you can use vectors together with functions like unique() to achieve the same behavior.


Make a list and a vector 

```{r}

my_list_own <- list("chai latte", "hot cocoa", "cappuccino")
  
my_namedlist_own <- list(sport = "climbing", age = 22)

my_vector_own <- c(3, 6, 7, 10, 11)

```


## Accessing elements 
Now we would like to access elements in our container variables, this is almost as in python. 
Vectors → accessed with [ ] using numeric positions.
Lists → accessed with [ ] for sublists, [[ ]] for elements, and $ for named elements.

Examples:
```{r}

# Vectors

numbers <- c(10, 20, 30, 40)

numbers[1]       # Access the first element → 10
numbers[c(2, 4)] # Access the 2nd and 4th elements → 20, 40

#When passing multiple numbers in r we usually use a vector like here c() to specify the numbers we need. 

# Lists / named lists 
my_list <- list(name = "Johanne", age = 25, scores = c(90, 85, 88))

my_list[1]      # Returns a sublist containing the first element → list(name = "Johanne")
my_list[[1]]    # Returns the actual element → "Johanne"
my_list$age    # Access by age → 25
my_list[[3]]    # Access the scores vector → c(90, 85, 88)

```

Excercise - access some elements in your lists and vector
```{r}

# for the vector: 

my_vector_own[3]

my_vector_own[c(4,5)]

# for the list:

my_namedlist_own[2]

my_namedlist_own$sport

```


## Condtional statements
In r we have to put the condition into a parenthesis () and wrap the whole statement in curly brackets {}

```{r}

# The if statement

x <- 5
if (x > 0) {
  print("x is positive")
}

```

```{r}

# The else statement

x <- -3

if (x > 0) {
  print("x is positive")
} else {  # Here there is a new set of curly brackets because there is a new condition
  print("x is not positive")
}

```

```{r}

# The else if (elif) statement

x <- 0

if (x > 0) {
  print("x is positive")
} else if (x < 0) {
  print("x is negative")
} else {
  print("x is zero")
}

```

Exercise: translate the following conditional statement from python to R

x = 5
if x > 5:
    print("x is greater than 5")
elif x == 5:
    print("x is equal to 5")
else:
    print("x is less than 5")
    
```{r contional statement}

x <- 5

if (x > 5) {
  print("x is greater than 5")
} else if (x == 5) {
  print("x is equal to 5")
} else {
  print("x is less than 5")
}

```

*logical orerators*



## Loops 
The syntax for loops is the same as for conditionals :)

Maybe you remember that python was indexed from 0 - r is not. Here we start at one.
What will this code chunck print?

```{r}

# a for loop

for (i in 1:5) {
  print(i)
}

# It will print 1, 2, 3, 4, 5

```


```{r}

# a while loop 

x <- 1
while (x <= 5) {
  print(x)
  x <- x + 1
}

# It will print 1, 2, 3, 4, 5

```

Excercise: translate the following loop from python to R

fruits = ["apple", "banana", "cherry"]
for fruit in fruits:
    print(fruit)


```{r}

fruits <- list("apple", "banana", "cherry")

for (fruit in fruits) {
  print(fruit)
}

```


## Functions
In r we have to specify a function using the function called function(). What?? I know, it's sort of weird but hang in. 

So the syntax is like this:

function_name <- function(arguments) {
  # code to execute
  return(value) # is we would like to return something
}

```{r}

add_numbers <- function(a, b) {
  sum <- a + b
  return(sum)
}

add_numbers(3, 5)

```

Exercise: translate the following function into R

# celsius to fahrenheit converter
def celcius_to_fahrenheit(celcius):
    F = (celcius * 9/5) +32
    print(f"{celcius} degrees celcius is the equivalent of {F} degress fahrenheit")
    return F

celcius_to_fahrenheit(0)

```{r}

celcius_to_fahrenheit <- function(celcius) {
  F <- (celcius * 9/5) + 32
  print(paste0(celcius," degrees celcius is the equivalent of ", F, " degrees fahrenheit"))
  return(F)
}

celcius_to_fahrenheit(32)

```



## Data Organisation
We will do a bit of data mangling as well using the same dataset, which is built in to R (smart).


```{r}

# Loading the data 

data(iris)  # load the dataset
class(iris)

head(iris)  # view the first few rows (like Python's head())

```

Try to click on the 'iris' dataset in the enviroment in the top right corner to actually inspect the dataset. 


*Accessing elements*
Acessing elements can be done in different ways, just like accessing values in container variables.
First, We access _Columns_
```{r accessing columns}

# Using the $ operator
iris$Sepal.Length

# Try to access another column this way
iris$Petal.Length

iris$Species

```
Can you figure out what the numbers in the right side stands for?

```{r accessing columns}

# Using column index
iris[, 1]

# Try to access another column this way
iris[, 3]

iris[, 5]

```


```{r accessing columns}

# Using column name in brackets
iris[ , "Sepal.Length"]

# Try to access another column this way
iris[ , "Petal.Width"]

iris[ , "Sepal.Width"]

```

Then, We access rows. And we do that by using numerical indicies (as they dont have names only numbers)
```{r}

# First row
iris[1, ]

# First three rows
iris[1:3, ]

```

If We want to access both rows and columns we can do that. Remember rows comes first: iris[rows, columns]

Excercise: access both rows and columns
```{r}

# Try to extract rows 1-3 and  columns 1-2
iris[1:3, 1:2]

# What about a single element: row 3, column 1
iris[3, 1]

# You can also use column name for a specific row. Try to get the same element as above
iris[3, "Sepal.Length"]


```

*Subsetting*
If you want to subset a part of the data you can do that using the subset() function and pass it a condition

```{r}

#Example
subset(iris, Sepal.Length > 5) # Select rows where Sepal.Length > 5


# If you only want specific data for this subset and not the whole dataframe you can select which columns you want. 
subset(iris, Sepal.Length > 5, select = c(Sepal.Length, Species))


```

You can also find the max and median using the functions max() and median(). In R we wrap the function around what we want to find e.g. max(iris$Petal.Length) where we did .max() at the end in python. 

```{r}

# find the max and median of a column

max <- max(iris$Petal.Length)
  
median <- median(iris$Petal.Length)

print(max)
print(median)

```


*Sorting* 
You can sort a vector or a column of a data frame using sort() for a vector or order() for a column in a dataframe.

```{r}

# Sort a data frame by a column
iris_sorted <- iris[order(iris$Sepal.Length), ]  # ascending by Sepal.Length

# Descending add a ,decreasing = TRUE inside the parenthesis
iris_sorted_desc <- iris[order(iris$Sepal.Length, decreasing = TRUE), ]

```

*Flipping/reversing*
To reverse a vector or the order of rows in a data frame, use rev()

```{r}

# Reverse rows in a data frame
iris_reversed <- iris[rev(1:nrow(iris)), ]

```



# Find five flaws 
In this exercise, you will work with the iris data set. Your goal is to find the 5 flaws in two exercises, all flaws are syntactical errors:


Exercise 1
Subset the data set to include only rows where Sepal.Length is greater than the median, Petal.Width is less than 1.5, and Species is "setosa".
Calculate the maximum and minimum of Sepal.Length in this subset and lastly calculate the mean of Petal.Length in this subset.

```{r} 

#Make comments about the flaws so you remember if you stumble upon them again

median_sepal <- median(iris$Sepal.Length) # median should be spelled with lower case letters.

subset_flowers <- subset(iris, Sepal.Length > median_sepal & Petal.Width < 1.5 & Species == "setosa") # we only need 1x "&" and not double "&&"

max_length <- max(subset_flowers$Sepal.Length) # brackets should always be closed off

min_length <- min(subset_flowers$Sepal.Length) # call function correctly - it's min() and not minimum

mean_petals <- mean(subset_flowers$Petal.Length) # data set should be called first and then "$" to select column

max_length
min_length
mean_petals

# Since no rows fulfills the requirements, there are no max, min and mean to find within this subset.

```

Exercise 2 
In this exercise, you will write and use a function to filter the iris data set. The function should:

Take a species name and a minimum petal length as input.

Return all rows where Species matches the input and Petal.Length is greater than the minimum.

Then, loop over all unique species in the dataset and print the number of flowers returned for each.

```{r}

filter_species <- function(species_name, min_petal) {
  filtered <- subset(iris, Species == species_name & Petal.Length > min_petal)
  return(filtered)
}

for (s in unique(iris$Species)) {   
  flowers <- filter_species(s, 1.5)
  print(paste("Number of flowers for", s, "=", nrow(flowers)))
}

filter_species("virginica", 6.1)

```

Juhuu - now you finished this translation markdown, so now you are super ready to write some r-code!!
