---
title: "Portfolio_2"
output: html_document
date: "05/10/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading packages}

# Install pacman first (only once):
install.packages("pacman")

# Load and install packages:
pacman::p_load(ggplot2, dplyr, tidyr, tidyverse, ggrepel, dslabs) # Installs the library if needed + imports it in one step

```

```{r loading dataset}

divorce <- dslabs::divorce_margarine

```


# Correlation and Regression: 

## PART 1: 

Use the ‘divorce_margarine’ dataset from the ‘dslabs’ package to explore the relationship
between margarine consumption and divorce rates in Maine.

Visualize the data, make a correlation test. Report the correlation coefficient and p-value in APA format, and briefly discuss the results in a practical way.

### Understand Data:

First, we have a quick look at the data:

```{r}

head(divorce)

```


Our data set consists of 10 observation of 3 variables:
- **divorce_rate_maine**
- **margarine_consumption_per_capita**
- **year**

The focus will be exploring the relationship between margarine consumption and divorce rates in Maine.

### Explore Distributions:

First we'll plot a histogram for each of the two variables: 

```{r}

# Margarine consumption:
ggplot(divorce, aes(x = margarine_consumption_per_capita)) + 
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "lightblue", color = "black") + 
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Margarine Consumption per Capita", x = "Margarine Consumption", y = "Density")

# Divorce rates: 
ggplot(divorce, aes(x = divorce_rate_maine)) +
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Divorce Rates in Maine", x = "Divorce Rates", y = "Density")

```
Both of the histograms don't look completely normal, as they have a bit of a skew on the left side (positive skew).

We can also have a look at the QQ-plots for each variable: 

```{r}

# Margarine consumption:
ggplot(divorce, aes(sample = margarine_consumption_per_capita)) + 
  stat_qq(color = "blue") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Margarine Consumption per Capita")

# Divorce rates: 
ggplot(divorce, aes(sample = divorce_rate_maine)) +
  stat_qq(color = "darkgreen") +
  stat_qq_line(color = "red") +
  labs(title = "Q-Q Plot: Divorce Rates in Maine")

```
Looking at the QQ-plots for each variable, the data now appears to look more normal, as the data follows the line nicely and doesn't deviate much. 

Lastly, we'll try running a Shapiro-Wilk test to further investigate the normality of the data: 

```{r}

shapiro.test(divorce$margarine_consumption_per_capita)
shapiro.test(divorce$divorce_rate_maine)

```
Both tests show p > 0.05, suggesting that the data can be treated as normal. Margarine consumption has a p-value of .2503, while divorce rates has a p-value of .07916 

These results together with the plots suggests that we can treat the data as normal. Thus, no transformations will be done. 

### Standardize Variables:

Both variables will be standardized in order to sufficiently compare the units of measurements. 

```{r}

scaled_margarine <- scale(divorce$margarine_consumption_per_capita)

scaled_divorce_rates <- scale(divorce$divorce_rate_maine)

head(scaled_margarine)

head(scaled_divorce_rates)

```
### Check Assumptions for Correlation:

In order to do correlation, we need to check assumptions for a parametric test: 
- **Normality of Distributed Data**
- **Homogeneity of Variance** 
- **Interval Data**
- **Data are Independent**

Given the above assumption and that we have no extreme outliers, we can assume the data is (approximately) normally distributed. Thus, we can proceed with a parametric test such as Pearson's correlation:

```{r}

cor.test(scaled_margarine, scaled_divorce_rates, method = "pearson")

```
Margarine consumption was found to be significantly correlated with divorce rates, r(8) = .9926, p = .0000000133, 95% CI [.9677, .9983]

Since the p-value < 0.05, there's a statistical significance. Also, since the correlation is very close to 1, there's a strong relationship between the variable. 

These results suggests that there's a high correlaton between the consumption of margarine and divorce rates in Maine. 


## PART 2:

Work with the ‘GSSvocab’ dataset from the ‘car’ package, focusing only on the year 1978
(subset that year and exclude missing values). Investigate how vocabulary test scores (vocab)
are related to education (educ). Present the relationship in a plot, fit a regression model, and
report results in APA format (coefficients, p-values, and R2), followed by a short practical
interpretation.

Extend the analysis by making a new model, where you include whether a person is native-
born (nativeBorn) as a predictor. Again, visualise and model the relationship.
Consider whether education and native-born status interact in predicting vocabulary, and if
so, fit an interaction model. For each model, report results in APA format and explain the
practical meaning.

Finally, compare the models and discuss which one performs best based on p-values and R2,
as well as the real-world interpretability of the results

### Regression Model 1: 

```{r}

# First, we'll download the package and load in the data set:

pacman::p_load(carData)

GSSvocab <- carData::GSSvocab

```

We'll start by subsetting the data, so we have only the year 1978 and no missing values: 

```{r}

# Subset year 1978: 

GSSvocab_1978 <- subset(GSSvocab, year == 1978)

# Remove missing values: 

GSSvocab_1978 <- drop_na(GSSvocab_1978)

```

### Regression Model 2:


### Comparing the Models:


