---
title: "Portfolio_2"
output: html_document
date: "05/10/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading packages}

# Install pacman first (only once):
install.packages("pacman")

# Load and install packages:
pacman::p_load(ggplot2, dplyr, tidyr, tidyverse, ggrepel, dslabs) #So it installs the library if needed + imports it in one step.

```

```{r loading dataset}

divorce <- dslabs::divorce_margarine

```


# Correlation and Regression: 

## PART 1: 

Use the ‘divorce_margarine’ dataset from the ‘dslabs’ package to explore the relationship
between margarine consumption and divorce rates in Maine.

Visualize the data, make a correlation test. Report the correlation coefficient and p-value in APA format, and briefly discuss the results in a practical way.

### Understand Data:

```{r}

head(divorce)

```


Our data set consists of 10 observation of 3 variables:
- **divorce_rate_maine**
- **margarine_consumption_per_capita**
- **year**

The focus will be exploring the relationship between margarine consumption and divorce rates in Maine.

### Explore Distributions:

First we'll plot a histogram for each of the two variables: 

```{r}

# Margarine consumption:
ggplot(divorce, aes(x = margarine_consumption_per_capita)) + 
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "lightblue", color = "black") + 
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Margarine Consumption per Capita", x = "Margarine Consumption", y = "Density")

# Divorce rates: 
ggplot(divorce, aes(x = divorce_rate_maine)) +
  geom_histogram(aes(y = after_stat(density)), bins = 20, fill = "lightgreen", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Divorce Rates in Maine", x = "Divorce Rates", y = "Density")

```



## PART 2:

Work with the ‘GSSvocab’ dataset from the ‘car’ package, focusing only on the year 1978
(subset that year and exclude missing values). Investigate how vocabulary test scores (vocab)
are related to education (educ). Present the relationship in a plot, fit a regression model, and
report results in APA format (coefficients, p-values, and R2), followed by a short practical
interpretation.

Extend the analysis by making a new model, where you include whether a person is native-
born (nativeBorn) as a predictor. Again, visualise and model the relationship.
Consider whether education and native-born status interact in predicting vocabulary, and if
so, fit an interaction model. For each model, report results in APA format and explain the
practical meaning.

Finally, compare the models and discuss which one performs best based on p-values and R2,
as well as the real-world interpretability of the results


